\begin{thebibliography}{14}
\newcommand{\enquote}[1]{``#1''}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{\texttt{#1}}
\providecommand{\urlprefix}{URL }
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
  \urlstyle{rm}\Url}\fi
\providecommand{\eprint}[2][]{\url{#2}}

\bibitem[{Allaire \emph{et~al.}(2018)Allaire, Eddelbuettel, and
  Fran\c{c}ois}]{CRAN:Rcpp:Attributes}
Allaire JJ, Eddelbuettel D, Fran\c{c}ois R (2018).
\newblock \emph{{Rcpp} Attributes}.
\newblock Vignette included in R package Rcpp,
  \urlprefix\url{http://CRAN.R-Project.org/package=Rcpp}.

\bibitem[{Ametrano \emph{et~al.}(2018)Ametrano, Ballabio, Bianchetti,
  C\'{e}sar\'{e}, Eddelbuettel, Firth, Jean, Kenyon, Lichters, Marchioro,
  Spanderen, and Wang}]{Ametrano+Ballabio+et+al:QuantLib}
Ametrano F, Ballabio L, Bianchetti M, C\'{e}sar\'{e} ND, Eddelbuettel D, Firth
  N, Jean N, Kenyon C, Lichters R, Marchioro M, Spanderen K, Wang J (2018).
\newblock \enquote{QuantLib: A free/open-source library for quantitative
  finance.}
\newblock Version 1.13., \urlprefix\url{http://quantlib.org/}.

\bibitem[{Box and Muller(1958)}]{Box+Muller:1958}
Box GEP, Muller ME (1958).
\newblock \enquote{A Note on the Generation of Random Normal Deviates.}
\newblock \emph{Annals of Mathematical Statistics}, \textbf{29}(2), 610--611.

\bibitem[{Brown \emph{et~al.}(2013)Brown, Eddelbuettel, and
  Bauer}]{Brown+Eddelbuettel+Bauer:DieHarder}
Brown RG, Eddelbuettel D, Bauer D (2013).
\newblock \enquote{Dieharder: A Random Number Test Suite.}
\newblock Open Source software library, under development,
  \urlprefix\url{http://www.phy.duke.edu/~rgb/General/dieharder.php}.

\bibitem[{Cottrell and Lucchetti(2018)}]{Cottrell+Lucchetti:Gretl}
Cottrell A, Lucchetti R (2018).
\newblock \emph{Gretl User's Guide}.
\newblock Version 2018a, \urlprefix\url{http://gretl.sourceforge.net/}.

\bibitem[{Eddelbuettel(2013)}]{Eddelbuettel:2013:Rcpp}
Eddelbuettel D (2013).
\newblock \emph{Seamless R and C++ Integration with Rcpp}.
\newblock Use R! Springer, New York.
\newblock ISBN 978-1-4614-6867-7.

\bibitem[{Eddelbuettel and Fran\c{c}ois(2015)}]{CRAN:Rcpp}
Eddelbuettel D, Fran\c{c}ois R (2015).
\newblock \emph{{Rcpp}: Seamless {R} and {C++} Integration}.
\newblock R package version 0.12.0,
  \urlprefix\url{http://CRAN.R-Project.org/package=Rcpp}.

\bibitem[{Eddelbuettel and Balamuta(2018)}]{TAS:Rcpp}
Eddelbuettel D, Balamuta JJ (2018).
\newblock \enquote{Extending R with C++: A Brief Introduction to Rcpp.}
\newblock \emph{The American Statistician}, \textbf{72:1}.
\newblock \url{10.1080/00031305.2017.1375990}.
  
\bibitem[{Galassi \emph{et~al.}(2013)Galassi, Davies, Theiler, Gough, Jungman,
  Alken, Booth, and Rossi}]{GSL}
Galassi M, Davies J, Theiler J, Gough B, Jungman G, Alken P, Booth M, Rossi F
  (2013).
\newblock \emph{{GNU} {S}cientific {L}ibrary {R}eference {M}anual}, 3rd
  edition.
\newblock Version 1.16. {ISBN} 0954612078,
  \urlprefix\url{http://www.gnu.org/software/gsl}.

\bibitem[{Leong \emph{et~al.}(2005)Leong, Zhang, Lee, Luk, and
  Villasenor}]{Leong+Zhang+Lee+Luk+Villasenor:2005}
Leong PHW, Zhang G, Lee DU, Luk W, Villasenor J (2005).
\newblock \enquote{A Comment on the Implementation of the Ziggurat Method.}
\newblock \emph{Journal of Statistical Software}, \textbf{12}(7), 1--4.
\newblock ISSN 1548-7660.
\newblock \urlprefix\url{http://www.jstatsoft.org/v12/i07}.

\bibitem[{Marsaglia and Tsang(2000)}]{Marsaglia+Tsang:2000}
Marsaglia G, Tsang WW (2000).
\newblock \enquote{The Ziggurat Method for Generating Random Variables.}
\newblock \emph{Journal of Statistical Software}, \textbf{5}(8), 1--7.
\newblock ISSN 1548-7660.
\newblock \urlprefix\url{http://www.jstatsoft.org/v05/i08}.

\bibitem[{Matsumoto and Nishimura(1998)}]{Matsumoto+Nishimura:1998}
Matsumoto M, Nishimura T (1998).
\newblock \enquote{Mersenne Twister: A 623-dimensionally Equidistributed
  Uniform Pseudo-random Number Generator.}
\newblock \emph{ACM Transactions on Modeling and Computer Simulation (TOMACS)},
  \textbf{8}(1), 3--30.
\newblock ISSN 1049-3301.
\newblock \doi{10.1145/272991.272995}.
\newblock \urlprefix\url{http://doi.acm.org/10.1145/272991.272995}.

\bibitem[{Thomas \emph{et~al.}(2007)Thomas, Luk, Leong, and
  Villasenor}]{Thomas+Luk+Leong+Villasenor:2007}
Thomas DB, Luk W, Leong PHW, Villasenor JD (2007).
\newblock \enquote{Gaussian random number generators.}
\newblock \emph{ACM Computing Surveys}, \textbf{39}(4), 11:1--11:38.

\bibitem[{Voss(2011)}]{Voss:2011}
Voss J (2011).
\newblock \enquote{The Ziggurat Method for Generating Gaussian Random Numbers.}
\newblock Webpage at \url{http://www.seehuhn.de/pages/ziggurat}.

\bibitem[{Yalta and Schreiber(2012)}]{Yalta+Schreiber:2012}
Yalta AT, Schreiber S (2012).
\newblock \enquote{Random Number Generation in gretl.}
\newblock \emph{Journal of Statistical Software, Code Snippets},
  \textbf{50}(1), 1--13.
\newblock ISSN 1548-7660.
\newblock \urlprefix\url{http://www.jstatsoft.org/v50/c01}.

\end{thebibliography}
